Biodiversity improves tropical forest resilience, shifting from complementarity to selection effect

### 2018-11-15 draft.Rmd
Draft 0.1.9008 Fifth version (IM references)

---
title: Biodiversity improves tropical forest resilience, shifting from complementarity to selection effect
author: "Sylvain Schmitt, Isabelle Maréchaux, Jerome Chave, Fabian Fischer, Camille Piponiot, Stéphane Traissac & Bruno Hérault"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
    keep_tex: true
  bookdown::word_document2: default
  github_document: default
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

Resilience concept has met numerous definitions, but resilience in ecology can encompass both ecosystem ability to adsorb and recover from disturbance, resistance and recovery respectively [@Gunderson2010]. Both exogenous (e.g. climate, disturbance features) and endogenous (e.g. forest structure and composition) factors can drive ecosystem recovery after disturbance. Among the former, logging intensity has been reported to strongly control and increase time-to-recovery of post-logging tree carbon, subsequent animal biodiversity loss, and change in tree species composition [@Burivalova2014; @DeAvila2015; @Rutishauser2015]. Among endogenous drivers, forest pre-disturbance composition has been identified as a key driver of forest recovery [@Herault2018], as different species respond differently to perturbations depending on their ecological strategies and functional properties [@herault_functional_2011]. Beyond species identity, tree community diversity may also foster forest resilience to perturbations [@Cadotte2011; @Ives2007; @Loreau2013]. Finally, climate change effect on tropical areas results in a significant risk of precipitation decrease [@Davidson2012], whereas intense drought events has been shown to result in huge mortality rates and biomass loss [@Phillips2009].

@@

The effect of biodiversity on ecosystem functioning has been intensely studied over the last decades, through theoretical work [@Loreau2010], experiments [@Cardinale2009], or observations along natural gradient of biodiversity [@Chisholm2013; @Liang2016; @Paquette2011; @Poorter2017]. However, experiments have mostly been conducted on grasslands [@Hector2005], while studies relying on forest stand inventories often lack repeatability [@Schnitzer2016], rarely include disturbed forests [@Sist2015], and monitoring temporal coverage remains too low to assess long-term effect as trees are long-lived organisms [@Herault2018]. Forest simulators have thus proven useful to investigate the effect of biodiversity on forest ecosystem functioning through virtual biodiversity experiments [@Morin2011; @Morin2014; @Sakschewski2016] However, the problem of biodiversity representation beyond a few plant functional types in models of tropical forests have long been an impediment to apply this approach to this species-rich ecosystem [@Kazmierczak2014; @Kohler1998; @Purves2008].

@@

We used a simulation approach to assess the effect of species richness and functional composition on long-term forest recovery from disturbances of different levels. Generally, a disturbance can be characterized by a loss of individuals at a specific time. Individual loss can be more or less targeted on individual characteristics (i.e. species, diameter...) and time can be cyclic or not. To have an overview of disturbance effect without confounding effect of disturbance type specificity, we used untargeted disturbance happening only once. We manipulated species number and identity across a large number of simulations, and measured the biodiversity net effect on simulated forest recovery after disturbance. In order to further investigate the mechanisms underlying such potential effect of biodiversity, we partitioned it into complementarity and selection effects. Based on the general hypothesis of a positive relationships between biodiversity and productivity [@Liang2016], we hypothesized that more diverse forest will recover quicker to a disturbance event due to an increased productivity.

@@

A new disturbance module was developed for this study. In order to have a general overview of disturbance effect on forest recovery, we used untargeted disturbance happening only once to be free of confounding factors due to disturbance type specificity. At a given iteration $disturb_{iter}$, individual trees are randomly picked following a uniform law and without any individuals nor species targeting. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined removed basal area ($disturb_{intensity}$).

@@

In the limit of the model, we were able to show that functional diversity improved tropical forest recovery from disturbance. More particularly, functional diversity and evenness are key components of diversity in forest recovery trajectory after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in the early times before more productive species dominate the forest and insure late recovery. We thus advocate the view that both diversity theory (with niche complementarity and facilitation) and mass ratio theory (represented by selection) are implied in ecosystem recovery but with relative importance depending on the spatial grain and time since the disturbance event.

@@

Our results validated the hypothesis of a significant positive relationship between forest recovery and functional diversity and evenness. More particularly, functional diversity [volume of the functional space occupied by the community, @villeger_new_2008] seems a major aspect of recovery if it's strengthened by a high functional evenness [regularity of the distribution of abundance in this volume, @villeger_new_2008]. Indeed, high functional diversity combined to high evenness avoid hegemonic effect of a few dominant species. Our results thus confirms both the review of @Diaz2001,advocating for an under-evaluated importance of plant functional diversity in ecosystem processes, and @Zhang2012 who highlighted the role of evenness in productivity and resilience of terrestrial ecosystems. Finally, species richness is not directly increasing resilience, thus supporting the idea that species diversity is often an inadequate surrogate of functional diversity [@Diaz2001; but see @Poorter2015]. Still, increased species richness will increase chance for high functional diversity through the sampling effect [@Loreau1998]. An higher sampling of regional species pool will allow a greater chance to pick more functionally diverse species.

@@

Our results follows several studies advocating for multiple mechanisms through which functional diversity affect ecosystem functions. Complementarity effect encompass niche complementarity theory (or diversity theory), which has been shown to indirectly improve forest biomass and coarse wood productivity in subtropical forest [@Chiang2016]. Similarly, complementarity has been shown of primary importance in grassland functioning [@Cardinale2007; @Isbell2011]. On the other hand, selection effect can be related to the mass ratio hypothesis, which proposes that ecosystem functions are regulated by dominant species of the community [@Grime1998]. And mass ratio theory has been shown to primarily influence ecosystem functions with biomass accumulation (aboveground and shoots), productivity (coarse wood or aboveground), water use, and light interception for both grassland and subtropical forests [@Chiang2016; @Mokany2008]. Consequently, concerning mechanisms responsible for the positive effect of functional diversity on tropical forest functions after disturbance, we also advocate for the interplay of both niche complementarity and mass ratio theories, depending on time since the disturbance event and the spatial grain.

Finally, our findings confirm results from experimental studies. @Williams2017 also found experimentally an over yielding due to complementarity effect based on light and spatial complementarity with tree crown of young trees. In addition, @Tobner2016 studied 4-years stand in low diverse forests of Canada. They also found that selection effect was greater than complementarity in most of the cases, promoting the idea that mass ratio theory is the primary mechanisms involved in functional diversity effect on ecosystem functions [@Mokany2008].

@@

Our results advocate for an important interplay of diversity and dominant species in forest resilience depending on the recovery stage, similarly to findings of @Sakschewski2016. Moreover, we studied a generic form of disturbance, and our findings might be modulated depending on the disturbance type. For instance, selective logging will induce both species targeting diminishing selection effect if dominant species are chosen and cyclic perturbation, increasing the role of early recovery where complementarity effect play the most. Consequently, we could hypothesize that biodiversity will be more of primary importance regarding selective logging compared to less targeted disturbances with longer cycles such as drought, fires and hurricanes.

>>>>

---
title: Biodiversity improves tropical forest resilience, shifting from complementarity to selection effect
author: "Sylvain Schmitt, Isabelle Maréchaux, Jerome Chave, Fabian Fischer, Camille Piponiot, Stéphane Traissac & Bruno Hérault"
date: '`r Sys.Date()`'
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
  bookdown::word_document2: default
  github_document: default
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
    keep_tex: true
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

Resilience concept has met numerous definitions, but resilience in ecology can encompass both ecosystem ability to adsorb and recover from disturbance, resistance and recovery respectively [@Gunderson2010]. Both exogenous (e.g. climate, disturbance features) and endogenous (e.g. forest structure and composition) factors can drive ecosystem recovery after disturbance. Among the former, logging intensity has been reported to strongly control and increase time-to-recovery of post-logging tree carbon, subsequent animal biodiversity loss, and change in tree species composition [@Burivalova2014; @DeAvila2015; @Rutishauser2015]. Among endogenous drivers, forest pre-disturbance composition has been identified as a key driver of forest recovery [@Herault2018], as different species respond differently to perturbations depending on their ecological strategies and functional properties [@herault_functional_2011]. Beyond species identity, tree community diversity may also foster forest resilience to perturbations [@DeAvila2018; @Cadotte2011; @Ives2007; @Loreau2013]. Finally, climate change effect on tropical areas results in a significant risk of precipitation decrease [@Davidson2012], whereas intense drought events has been shown to result in huge mortality rates and biomass loss [@Phillips2009].

@@

The effect of biodiversity on ecosystem functioning has been intensely studied over the last decades, through theoretical work [@Loreau2010], experiments [@Cardinale2007; @Cardinale2009; @Niklaus2017a; @Huang2018], or observations along natural gradient of biodiversity [@Chisholm2013; @Liang2016; @Paquette2011; @Poorter2017]. Whereas several studies advocate only for the importance of few dominant species through selection effect to regulate ecosystem functionning in an old tree diversity experiment [@Bauters2015], in wet or dry natural forests [@Finegan2015; @Prado-Junior2016], or in human-modified forests [@DeAvila2018; @Lohbeck2016]. Numerous authors highlighted diversity role through complementarity, alone or in addition to selection. @Morin2014a identified by simulation an increases of ecosystem stability with species richness. @Fargione2007 observed a shift from sampling to complementarity effect, when @Huang2018 observed an increased complementarity with a null selection effect in forest establishments experiments. And @Cardinale2007; @Poorter2017; @VanderSande2017a reviewed grasslands and forests experiments and study sites and highlighted the role of both complementarity and selection for ecosystem biomass stock and dynamic. However, experiments have mostly been conducted on grasslands [@Hector2005], while studies relying on forest stand inventories often lack repeatability [@Poorter2017; @Schnitzer2016], rarely include disturbed forests [@Sist2015], and monitoring temporal coverage remains too low to assess long-term effect as trees are long-lived organisms [@Herault2018]. Forest simulators have thus proven useful to investigate the effect of biodiversity on forest ecosystem functioning through virtual biodiversity experiments [@Morin2011; @Morin2014a; @Sakschewski2016] However, the problem of biodiversity representation beyond a few plant functional types in models of tropical forests have long been an impediment to apply this approach to this species-rich ecosystem [@Kazmierczak2014; @Kohler1998; @Purves2008].

@@

We used a simulation approach to assess the effect of species richness and functional composition on long-term forest recovery from disturbances of different levels. A disturbance can be characterized by a loss of individuals, targeted or not (i.e. species, diameter...), at a specific time, cyclic or not. To have an overview of disturbance effect without confounding effect of disturbance specificities, we used untargeted disturbance happening only once. We manipulated species number and identity across a large number of simulations, and measured the biodiversity net effect on simulated forest recovery after disturbance. In order to further investigate the mechanisms underlying such potential effect of biodiversity, we partitioned it into complementarity and selection effects. Based on the general hypothesis of a positive relationships between biodiversity and productivity [@Liang2016], we hypothesized that more diverse forest will recover quicker to a disturbance event due to an increased productivity [but see @VanderSande2018].

@@

A new disturbance module was developed for this study. In order to have a general overview of disturbance effect on forest recovery, we used untargeted disturbance happening only once to be free of confounding factors due to disturbance type specificities. At a given iteration $disturb_{iter}$, individual trees are randomly picked following a uniform law and without any individuals nor species targeting. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined removed basal area ($disturb_{intensity}$).

@@

In the limit of the model, we were able to show that functional diversity improved tropical forest resilience from disturbance. More particularly, functional diversity and evenness are key components of diversity in forest recovery trajectory after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in the early times before more productive species dominate the forest and insure late recovery. We thus advocate the view that both diversity theory (with niche complementarity and facilitation) and mass ratio theory (represented by selection) are implied in ecosystem resilience but with relative importance depending on the spatial grain and time since the disturbance event.

@@

Our results validated the hypothesis of a significant positive relationship between forest recovery and functional diversity and evenness. More particularly, functional diversity [volume of the functional space occupied by the community, @villeger_new_2008] seems a major aspect of recovery if it's strengthened by a high functional evenness [regularity of the distribution of abundance in this volume, @villeger_new_2008]. Indeed, high functional diversity combined to high evenness avoid hegemonic effect of a few dominant species. Our results thus confirms both the review of @Daz2001a, advocating for an under-evaluated importance of plant functional diversity in ecosystem processes, and @Zhang2012 who highlighted the role of evenness in productivity and resilience of terrestrial ecosystems. Finally, species richness is not directly increasing resilience, thus supporting the idea that species diversity is often an inadequate surrogate of functional diversity [@Daz2001a; but see @Poorter2015], which might explain the absence of diversity effect on resilience observed by @VanderSande2018. Still, increased species richness will increase chance for high functional diversity through the sampling effect [@Loreau1998; @Morin2014a]. An higher sampling of regional species pool will allow a greater chance to pick more functionally diverse species.

@@

Our results follows several studies advocating for multiple mechanisms through which functional diversity affect ecosystem functions. Complementarity effect encompass niche complementarity theory (or diversity theory), which has been shown to improve forest biomass [@Poorter2017; @Huang2018] and coarse wood productivity [@Chiang2016] in tropical and subtropical forests. Similarly, complementarity has been shown of primary importance in grassland functioning [@Cardinale2007; @Fargione2007; @Isbell2011]. On the other hand, selection effect can be related to the mass ratio hypothesis, which proposes that ecosystem functions are regulated by dominant species of the community [@Grime1998]. And mass ratio theory has been shown to primarily influence ecosystem functions with biomass accumulation (aboveground and shoots), productivity (coarse wood or aboveground), water use, and light interception for both grassland and forests [@Finegan2015; @Lohbeck2016; @Prado-Junior2016; @Chiang2016; @Mokany2008]. Consequently, concerning mechanisms responsible for the positive effect of functional diversity on tropical forest functions after disturbance, we also advocate for the interplay of both niche complementarity and mass ratio theories, depending on time since the disturbance event and the spatial grain.

Finally, our findings confirm results from experimental studies. @Williams2017 also found experimentally an over yielding due to complementarity effect based on light and spatial complementarity with tree crown of young trees. In addition, @Tobner2016 studied 4-years stand in low diverse forests of Canada. They also found that selection effect was greater than complementarity in most of the cases, promoting the idea that mass ratio theory is the primary mechanisms involved in functional diversity effect on ecosystem functions [@Mokany2008]. Besides @Huang2018 highlighted increased complementarity over null selection effect resulting in higher productivity due to tree growth in an 8-years forest experiment of China.

@@

Our results advocate for an important interplay of diversity and dominant species in forest resilience depending on the recovery stage, similarly to findings of @Sakschewski2016. Moreover, we studied a generic form of disturbance, and our findings might be modulated depending on the disturbance specificities. For instance, selective logging will induce both species targeting diminishing selection effect if dominant species are chosen and cyclic perturbation, increasing the role of early recovery where complementarity effect play the most. Consequently, we could hypothesize that biodiversity will be more of primary importance regarding selective logging compared to less targeted disturbances with longer cycles such as drought, fires and hurricanes.
